---
title: "NYPD_Data"
output: pdf_document
date: "2024-10-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NYPD Historical Shooting Data Project 

For this project, I chose to analyze NYPD historic shooting data to understand more about how reported shootings and murders have changed over time in each of NYC's boroughs. The dataset includes data from 2006 - 2023. I chose to primarily look at data in what I would consider "public" places, so I have excluded data from unknown or unrecorded locations, as well as private residences. 

## Bias Identification

It is important to note for this project that there may be sources of bias both in the data collection and aggregation, and in my analysis itself. 

1. The data included only includes reported incidents. One can assume the actual numbers are higher than the NYPD database would reflect. 

2. I have personally excluded data that occurs at a private residence in order to examine the likelihood of the average person being shot or murdered in each of the boroughs over time. I was interested in examining if NYC has gotten more or less dangerous over time, but I have excluded the private location data under the assumption that these shootings are between parties that know eachother, rather than a random act of violence. This is again, an assumption that reflects my personal bias. 

## Setting Up 
Some graphics in this R Markdown document require XQuartz for MacOS to run properly. 

```{r get_data,  warning = FALSE, message = FALSE, results = 'hide'}
# Set Up Environment
library(tidyverse)
library(lubridate)
library(forecast)
library(mgcv)
library(ggplot2)

# Read the CSV data from the URL 
nyc_data <- read.csv('https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD') 

```


```{r clean_data,  warning = FALSE, message = FALSE, results = 'hide'}

# Remove columns 1, 5-8, and 17-21 
data_reduced <- nyc_data %>% select(-c(1, 5:8, 17:21))

# Cleaning up empty or null entries 
data_reduced <- data_reduced %>% mutate(across(everything(), ~replace_na(.x, 'UNKNOWN'))) %>% mutate(across(everything(), ~ifelse(.x == '', 'UNKNOWN', .x))) 
data_reduced <- data_reduced %>% mutate(across(everything(), ~ifelse(. == "(null)", "UNKNOWN", .)))

# Condensing Entries for Location to Public, Private or Unknown
locations = unique(data_reduced$LOCATION_DESC)
private <- c('MULTI DWELL - APT BUILD','PVT HOUSE')
unknown <- c('NONE','UNKNOWN')
tmp = setdiff(locations, private)
public = setdiff(tmp, unknown)

data_reduced$LOCATION_DESC <- ifelse(data_reduced$LOCATION_DESC %in% public, 'PUBLIC', data_reduced$LOCATION_DESC)
data_reduced$LOCATION_DESC <- ifelse(data_reduced$LOCATION_DESC %in% private, 'PRIVATE', data_reduced$LOCATION_DESC)
data_reduced$LOCATION_DESC <- ifelse(data_reduced$LOCATION_DESC %in% unknown, 'UNKNOWN', data_reduced$LOCATION_DESC)
```

```{r organize_data,  warning = FALSE, message = FALSE}

#Original Location Data
view(locations)
#Condensed Location Data
new_locations = unique(data_reduced$LOCATION_DESC)
view(new_locations)
```

##Visualizing Public Shootings by Borough
The below plots investigate the number of shootings in each borough, and then examine a subset of those shootings that were classified as murders. I chose to examine only the shootings and murders that occurred in public places, which was included in the dataset as the location description. 

```{r plots,  warning = FALSE, message = FALSE}
#Questions to Answer
# 1. Let's examine public shootings only, by borough over time 
data_reduced <- data_reduced %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE))
boro_monthly_data <- data_reduced %>% mutate(year_month = floor_date(OCCUR_DATE, "month")) %>% group_by(BORO, LOCATION_DESC, year_month, STATISTICAL_MURDER_FLAG) %>% summarize(CASES = n())
boro_monthly_data <- boro_monthly_data %>% rename(OCCUR_DATE = year_month)

boro_public_plot <- boro_monthly_data %>% filter(LOCATION_DESC == "PUBLIC") %>% ggplot(aes(x = OCCUR_DATE, y = CASES, color = BORO)) + geom_line() + labs(title = "Monthly Shootings Over Time by Borough (Public Locations Only)", x = "Date", y = "Number of Cases") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(boro_public_plot) 

# 2. Add murder flag to be true filter 
boro_murder_plot <- boro_monthly_data %>% filter(LOCATION_DESC == "PUBLIC", STATISTICAL_MURDER_FLAG == "true") %>% ggplot(aes(x = OCCUR_DATE, y = CASES, color = BORO)) + geom_line() + labs(title = "Monthly Murder Shootings Over Time by Borough (Public Locations Only)", x = "Date", y = "Number of Cases") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(boro_murder_plot) 

# 3. Reduce to yearly data 
yearly_data <- boro_monthly_data %>% mutate(YEAR = year(OCCUR_DATE))
boro_yearly_data <- yearly_data %>% group_by(YEAR, BORO, LOCATION_DESC, STATISTICAL_MURDER_FLAG) %>% summarise(CASES = sum(CASES))

#Plot Yearly Data
boro_yearly_public_plot <- boro_yearly_data %>% filter(LOCATION_DESC == "PUBLIC") %>% ggplot(aes(x = YEAR, y = CASES, color = BORO)) + geom_line() + labs(title = "Yearly Shootings Over Time by Borough (Public Locations Only)", x = "Date", y = "Number of Cases") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(boro_yearly_public_plot) 

# Add murder flag 
boro_yearly_murder_plot <- boro_yearly_data %>% filter(LOCATION_DESC == "PUBLIC", STATISTICAL_MURDER_FLAG == "true") %>% ggplot(aes(x = YEAR, y = CASES, color = BORO)) + geom_line() + labs(title = "Yearly Murder Shootings Over Time by Borough (Public Locations Only)", x = "Date", y = "Number of Cases") + theme_minimal() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
print(boro_yearly_murder_plot) 

#Examine Safest and Least Safe Months
filtered_data <- boro_monthly_data %>% filter(LOCATION_DESC == "PUBLIC") %>% group_by(OCCUR_DATE, BORO) %>% summarise(TOTAL_CASES = sum(CASES))
lowest_cases_per_boro <- filtered_data %>% group_by(BORO) %>% slice_min(TOTAL_CASES, n = 1) %>% ungroup() 
highest_cases_per_boro <- filtered_data %>% group_by(BORO) %>% slice_max(TOTAL_CASES, n = 1) %>% ungroup() 

year_filtered_data <- boro_yearly_data %>% filter(LOCATION_DESC == "PUBLIC") %>% group_by(YEAR, BORO) %>% summarise(TOTAL_CASES = sum(CASES))
lowest_year_per_boro <- year_filtered_data %>% group_by(BORO) %>% slice_min(TOTAL_CASES, n = 1) %>% ungroup() 
highest_year_per_boro <- year_filtered_data %>% group_by(BORO) %>% slice_max(TOTAL_CASES, n = 1) %>% ungroup() 
```

## Examining High and Low Records
The tables below show the highest and lowest shooting recorded for each borough. 

```{r tables}
print(lowest_year_per_boro)
print(highest_year_per_boro)
```

## Predicting Murders as a Function of Total Shootings
Here I use a linear model to examine the number of murders compared to the total shootings. The plots below use the same model, but plot against two different metrics, 1. The Overall Shootings and 2. The Non Murderous Shootings. 

```{r model,  warning = FALSE, message = FALSE}

#Add Linear Model to either public or private shootings
boro_murders_shootings <- boro_monthly_data %>% filter(LOCATION_DESC == "PUBLIC") %>% group_by(BORO, OCCUR_DATE) %>% summarise( MURDERS = sum(CASES[STATISTICAL_MURDER_FLAG == "true"]), SHOOTINGS = sum(CASES[STATISTICAL_MURDER_FLAG == "false"]), TOTAL = sum(MURDERS,SHOOTINGS))
bronx_murders_shootings <- boro_murders_shootings %>% filter(BORO == 'BRONX')
mod = lm(MURDERS ~ TOTAL, data = bronx_murders_shootings)
bronx_pred <- bronx_murders_shootings %>% mutate(pred = predict(mod))


bronx_pred %>% ggplot() + geom_line(aes(x = TOTAL, y = MURDERS), color = "blue") + geom_line(aes(x = TOTAL, y = pred), color = "red") + labs(title = "Proportion of Actual and Predicted Murders vs Total Shootings", x = "Total Shootings", y = "Number of Murders") + theme_minimal()

bronx_pred %>% ggplot() + geom_line(aes(x = SHOOTINGS, y = MURDERS), color = "blue") + geom_line(aes(x = SHOOTINGS, y = pred), color = "red") + labs(title = "Proportion of Actual and Predicted Murders vs Non Murderous Shootings", x = "Non Murderous Shootings", y = "Number of Murders") + theme_minimal()


```